Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-12 doc/invt.tex
invt.tex: edit abstract, introduction, appendix

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by constructing a bias potential that offsets
  the potential of mean force (PMF) on the fly.
  %
  The two, however, differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bin containing the current $z$,
  while metadynamics updates also
  several neighboring bins according to
  a Gaussian wave packet.
  %
  The error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a method of computing the optimal schedule
  for a general updating scheme
  involving multiple bins.
  %
  We show that the optimal schedule of metadynamics
  depends on the simulation length
  and the width of the Gaussian.
  %is solved from the equation of motion
  %of a free particle with a position-dependent mass.
  %
  Further,
  the single-bin updating scheme
  used in the WL algorithm
  belongs to a class of multiple-bin updating schemes
  optimal for asymptotic convergence,
  and the optimal schedules of these schemes
  are all given by the inverse-time formula.
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\log p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Normally, the distribution $p^*(z)$ is
limited in one or a few narrow $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the entire curve.
%
This solution can be clumsy in practice
and it often fails to converge quickly
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially flatten
the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%
This alleviates the above problem of local trapping
by forcing the system to travel more
deliberately along $z$,
which often represents
a direction of slow motion.
%
To achieve a flat distribution, however,
the method must be able to construct a
bias potential that exactly cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built up on-the-fly
during simulation.
%
These techniques employ updating loops
that discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
The main difference between the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian wave packet
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which brings about a side effect
of disrupting the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

The updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014},
therefore,
determines the precision of the final bias potential,
hence that of the PMF.
%
For the WL algorithm, it is known
that the optimal updating magnitude, $\ln f$,
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
and the number of bins
(with ``time'' being
the number of simulation steps).
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case.
%
It also predicts, for metadynamics,
a more algebraically complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, we show that
the single-bin updating scheme
used by the WL algorithm
can be generalized to class of multiple-bin updating schemes
that are partially optimal for infinitely long simulations,
and the inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsbin.pdf}
  }
  \caption{
    \label{fig:errsbin}
    Error, $E$, versus the
    relative updating magnitude,
    %proportionality constant,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the single-bin (WL) scheme.
    %
    The results have been averaged over $1000$ independent runs.
    %
    \note{The figure was produced by \texttt{doc/fig/errsbin.gp}.
      The data were produced by the script \texttt{data/invt2run.py},
      and saved under \texttt{data/singlebin}
      (see the \texttt{Makefile} there).
    }
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errnnbr.pdf}
  }
  \caption{
    \label{fig:errnnbr}
    Normalized error, $(T + t_0) \, E$,
    versus the %proportionality constant,
    relative updating magnitude,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the nearest-neighbor updating scheme with $\mu_1 = 0.24$.
    %
    The results have been averaged over $1000$ independent runs.
    \note{This figure was produced by \texttt{doc/fig/errnnbr.gp}.
      The data were produced by \texttt{data/invt2run.py},
      saved under \texttt{data/nb0.24}
      (see the \texttt{Makefile} there).
    }
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt},
    versus the optimized inverse-time schedule
    Eq. \eqref{eq:alpha_invtlambda}
    for the Gaussian updating scheme
    with $\sigma = 10$.
    \note{This figure was produced by \texttt{doc/fig/optacmp.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h}.
      The data were saved under the directory \texttt{data/opta}.
      See the \texttt{make} object \texttt{sigsingle}
      in the \texttt{Makefile} there.
    }
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    Bandpass updating schemes.
    (a) Optimal schedule for a periodic variable
    with $K = 3$,
    and that for a non-periodic variable
    with $K = 6$.
    (b) Updating kernels
    compared to a Gaussian with a similar width,
    $\sigma = 5$.
    The lines are a guide to the eyes.
    Inset. Updating matrix in the non-periodic case
    is free of the zigzaggedness shown in the kernel.
    \note{This figure was produced by \texttt{doc/fig/sinc.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h} for panel (a),
      and the functions \texttt{savewin()}
      and \texttt{savewinmat()}
      in \texttt{prog/invt.h} for panel (b),
      The data were saved under the directory \texttt{data/sinc}.
      See the \texttt{make} object \texttt{sincsingle}
      in the \texttt{Makefile} there.
    }
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the bandpass updating schemes.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
    %
    Here, we assumed $t_0 \equiv T/(e^{q(T)} - 1)$
    for the optimal value of $q(T)$.
    %
    The points were obtained from averaging results
    from 1000 independent simulations;
    the curves are theoretical predictions.
    %
    \note{This figure was produced by
      \texttt{doc/fig/errsigscan.gp}.
      The data were saved under
      \texttt{data/scan}.
    }
  }
\end{center}
\end{figure}

@@

As an alternative to the
optimal schedule from Eq. \eqref{eq:q_opt},
%
here we consider the inverse-time formula,
Eq. \eqref{eq:alpha_invtlambda},
with $\lambda$ optimized variationally.
%
This approach allows to obtain
an explicit formula for the error,
which can be tested against simulation.
%
%Several previous studies\cite{marsili2006, barducci2008, dickson2011}
%have been focused on analytically simpler schedules.
%%
%In a similar spirit,
%we consider here
%the inverse-time formula,
%Eq. \eqref{eq:alpha_invtlambda},
%%
%%
%We shall compute the error computed from
%this variationally optimized schedule
%and compare it with that from the optimal schedule.
%%
%We shall show that the optimal parameter, $\lambda$,
%is sensitive to the simulation time, $T$,
%which parallels the gradual shift of
%the instantaneous eigenvalue,
%$\lambda(t)$,
%as discussed in Sec. \ref{sec:optschedule}.

@@

Using Eq. \eqref{eq:alpha_invtlambda}
in Eq. \eqref{eq:error_asym1} yields
%
\begin{align}
\Err_A
&=
\frac{    1    }
     { T + t_0 }
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k^2 }
       {    2 \, \nu_k - 1   }
\left[
  1 - \left(
        \frac {     t_0 }
              { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
\right],
\label{eq:error_asym_invt}
\end{align}
%
where $\nu_k \equiv \lambda_k / \lambda$.
%
The error is asymptotically minimal
only if $2\,\nu_k > 1$ for every $k$.
\note{At long times, $T \to \infty$, we get
$$
\begin{aligned}
  \Err_A
  =
  \sum_{k = 2}^n
  \begin{dcases}
    \frac{    1    }
         { T + t_0 }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   2 \, \nu_k - 1    }
    &
    \mathrm{if \;} 2 \, \nu_k > 1,
    \\%[1em]
    %
    %
    \frac{    \Gamma_k    }
         { 4 \, (T + t_0) }
    \ln \frac{ T + t_0 }
             {   t_0   }
    &
    \mathrm{if \;} 2 \, \nu_k = 1,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \nu_k  - 1}  }
         { (T + t_0)^{ 2 \, \nu_k } }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   1 - 2 \, \nu_k    }
    &
    \mathrm{if \;} 2 \, \nu_k < 1.
  \end{dcases}
\end{aligned}
$$
%
The last two cases
are asymptotically suboptimal
as they decay slower than $1/T$.
}
%
In this case,
we can drop the second term
in the square brackets,
and identify the optimal $\lambda$ as
the smallest real root of
%
\begin{equation}
  \sum_{k = 2}^n
    \frac { \lambda_k - \lambda }
          { \left(2 - \lambda/ \lambda_k \right)^2 }
  = 0.
\notag
%\label{eq:optimal_lambda_approx}
\end{equation}
%
Since
$$
\frac{ \nu_k^2        }
     { 2 \, \nu_k - 1 }
=
\frac{ \lambda_k^2 }
     { \lambda \, (2 \, \lambda_k - \lambda) }
\ge 1,
$$
with the equality achieved only at $\lambda = \lambda_k$,
the optimal scheme must satisfy
$\lambda_2 = \dots = \lambda_n = \lambda$.
%
%Now a matrix with uniform eigenvalues
%is a multiple of the identity matrix,
%which corresponds to the single-bin updating scheme.
%
%The above result means that
%the single-bin updating scheme is best
%for the asymptotic convergence.
%
%This result is discussed in more details in
This again reflects the asymptotic optimality
of the single-bin scheme, as discussed in
Sec. \ref{sec:optWL}.

For a finite-length simulation,
we need to include the residual error, $\Err_R$.
%
By Eq. \eqref{eq:alpha_invtlambda}, we have
%
\begin{equation}
  q(T)
  =
  \frac{ 1 } { \lambda }
  \log\left(
    \frac{ T + t_0 } { t_0 }
  \right)
  ,
\notag
%\label{eq:qt_invtlambda}
\end{equation}
%
so that Eq. \eqref{eq:error_res1} becomes
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that
the system was initially equilibrated
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql},
as well as Eq. \eqref{eq:t0_sinc}.
%
The total error is
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \frac
  {
    \Gamma_k \, \nu_k \,
    \left[
      \nu_k
      +
      (\nu_k - 1)
      \left(
        \frac{ t_0 } { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
    \right]
  }
  {
    (T + t_0) \, (2 \, \nu_k - 1)
  }
  .
\label{eq:error_invt}
\end{align}
%
%where
%$$
%\Err^f_k
%\equiv
%\left\langle
%  y_k^2
%\right\rangle_{ \alpha(t) }
%=
%\frac{  \Gamma_k \, \nu_k   }
%     {    2 \, (T + t_0)    }
%,
%$$
%is the equilibrium, or saturated, error of mode $k$
%at the terminal updating magnitude
%of $\alpha(T) = 1/[\lambda (T + t_0)]$.

>>>>

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing
  \note{``incrementally building up''?}
  a bias potential to offset
  the potential of mean force (PMF).
  %
  The two differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bias potential of
  the bin containing the current $z$,
  while metadynamics updates also
  those of neighboring bins according to
  a Gaussian wave packet.
  %
  The error of the final bias potential,
  hence that of the PMF,
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a method of computing the optimal schedule
  for a general multiple-bin updating scheme.
  %
  We show that the optimal schedule of metadynamics
  depends on the simulation length
  and the width of the Gaussian.
  %is solved from the equation of motion
  %of a free particle with a position-dependent mass.
  %
  Further,
  the single-bin updating scheme
  used in the WL algorithm
  belongs to a class of multiple-bin bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  The optimal schedule of these schemes
  is given by the inverse-time formula,
  with a shifted origin of time such that
  the initial updating magnitude is half of
  the previous equilibrium value.
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the task is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\log p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Normally, the distribution $p^*(z)$ is
limited in one or a few narrow $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the entire curve.
%
This solution can be laborious % clumsy in practice
and it often fails to converge quickly
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum.
%for a long time.

A better solution is to artificially flatten
the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%
This alleviates the above problem of local trapping
by forcing the system to travel more
deliberately along $z$,
which often represents
a direction of slow motion.
%
To achieve a flat distribution, however,
the method must be able to construct a
bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is gradually built up
during simulation.
%
These techniques employ updating loops
to discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
The main difference between the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian wave packet
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which % brings about a side effect of disrupting
disrupts \note{``disturbs''?} the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

The updating schedule,
or the manner of reducing
the updating magnitude,
therefore,
determines the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal updating magnitude, $\ln f$,
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
and the number of bins
(with ``time'' being
the number of simulation steps).
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case,
and gives the optimal value of the origin of time.
%
It also predicts, for metadynamics,
a more complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, we generalize
the single-bin updating scheme
used by the WL algorithm
to a class of multiple-bin updating schemes
that are optimal for infinitely long simulations,
and the inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsbin.pdf}
  }
  \caption{
    \label{fig:errsbin}
    Error, $E$, versus the
    relative updating magnitude,
    %proportionality constant,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the single-bin (WL) scheme.
    %
    The results have been averaged over $1000$ independent runs.
    %
    \note{The figure was produced by \texttt{doc/fig/errsbin.gp}.
      The data were produced by the script \texttt{data/invt2run.py},
      and saved under \texttt{data/singlebin}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errnnbr.pdf}
  }
  \caption{
    \label{fig:errnnbr}
    Normalized error, $(T + t_0) \, E$,
    versus the %proportionality constant,
    relative updating magnitude,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the nearest-neighbor updating scheme with $\mu_1 = 0.24$.
    %
    The results have been averaged over $1000$ independent runs.
    %
    \note{This figure was produced by \texttt{doc/fig/errnnbr.gp}.
      The data were produced by \texttt{data/invt2run.py},
      saved under \texttt{data/nb0.24}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt},
    versus the optimized inverse-time schedule
    Eq. \eqref{eq:alpha_invtlambda}
    for the Gaussian updating scheme
    with $\sigma = 10$.
    %
    \note{This figure was produced by \texttt{doc/fig/optacmp.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h}.
      The data were saved under the directory \texttt{data/opta}.
      See the \texttt{make} object \texttt{sigsingle}
      in the \texttt{Makefile} there.
    }%
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    Bandpass updating schemes.
    (a) Optimal schedule for a periodic variable
    with $K = 3$,
    and that for a non-periodic variable
    with $K = 6$.
    (b) Updating kernels
    compared to a Gaussian with a similar width,
    $\sigma = 5$.
    The lines are a guide to the eyes.
    Inset. Updating matrix in the non-periodic case
    is free of the zigzaggedness shown in the kernel.
    %
    \note{This figure was produced by \texttt{doc/fig/sinc.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h} for panel (a),
      and the functions \texttt{savewin()}
      and \texttt{savewinmat()}
      in \texttt{prog/invt.h} for panel (b),
      The data were saved under the directory \texttt{data/sinc}.
      See the \texttt{make} object \texttt{sincsingle}
      in the \texttt{Makefile} there.
    }%
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the bandpass updating schemes.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
    %
    Here, we assumed $t_0 \equiv T/(e^{q(T)} - 1)$
    for the optimal value of $q(T)$.
    %
    The points were obtained from averaging results
    from 1000 independent simulations;
    the curves are theoretical predictions.
    %
    \note{This figure was produced by
      \texttt{doc/fig/errsigscan.gp}.
      The data were saved under
      \texttt{data/scan}.
    }%
  }
\end{center}
\end{figure}

@@

As an alternative to the
optimal schedule from Eq. \eqref{eq:q_opt},
%
here we consider the analytically simpler
inverse-time formula, Eq. \eqref{eq:alpha_invtlambda},
with $\lambda$ optimized variationally.
%
This approach allows us to compute the error explicitly.

@@

Using Eq. \eqref{eq:alpha_invtlambda}
in Eq. \eqref{eq:error_asym1} yields
%
\begin{align}
\Err_A
&=
\frac{    1    }
     { T + t_0 }
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k^2 }
       {    2 \, \nu_k - 1   }
\left[
  1 - \left(
        \frac {     t_0 }
              { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
\right],
\label{eq:error_asym_invt}
\end{align}
%
where $\nu_k \equiv \lambda_k / \lambda$.
%
The error is asymptotically minimal
only if $2\,\nu_k > 1$ for every $k$.
%
\note{At long times, $T \to \infty$, we get
$$
\begin{aligned}
  \Err_A
  =
  \sum_{k = 2}^n
  \begin{dcases}
    \frac{    1    }
         { T + t_0 }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   2 \, \nu_k - 1    }
    &
    \mathrm{if \;} 2 \, \nu_k > 1,
    \\%[1em]
    %
    %
    \frac{    \Gamma_k    }
         { 4 \, (T + t_0) }
    \ln \frac{ T + t_0 }
             {   t_0   }
    &
    \mathrm{if \;} 2 \, \nu_k = 1,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \nu_k  - 1}  }
         { (T + t_0)^{ 2 \, \nu_k } }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   1 - 2 \, \nu_k    }
    &
    \mathrm{if \;} 2 \, \nu_k < 1.
  \end{dcases}
\end{aligned}
$$
%
The last two cases
are asymptotically suboptimal
as they decay slower than $1/T$.
}%
%
In this case,
we can drop the second term
in the square brackets,
and identify the optimal $\lambda$ as
the smallest real root of
%
\begin{equation}
  \sum_{k = 2}^n
    \frac { \lambda_k - \lambda }
          { \left(2 - \lambda/ \lambda_k \right)^2 }
  = 0.
\notag
%\label{eq:optimal_lambda_approx}
\end{equation}
%
Since
$$
\frac{ \nu_k^2        }
     { 2 \, \nu_k - 1 }
=
\frac{ \lambda_k^2 }
     { \lambda \, (2 \, \lambda_k - \lambda) }
\ge 1
,
$$
with the equality achieved only at $\lambda = \lambda_k$,
the optimal scheme must satisfy
$\lambda_2 = \dots = \lambda_n$,
corresponding to the single-bin scheme
(cf. Sec. \ref{sec:optWL}).

For a finite-length simulation,
we need to include the residual error, $\Err_R$.
%
Assuming Eqs. \eqref{eq:error_res1} and \eqref{eq:t0_sinc},
we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\notag
%\label{eq:error_res_invt}
\end{equation}
%
with
%
\begin{equation}
  q(T)
  =
  \frac{ 1 } { \lambda }
  \log\left(
    \frac{ T + t_0 } { t_0 }
  \right)
  ,
\notag
%\label{eq:qt_invtlambda}
\end{equation}
%
from integrating Eq. \eqref{eq:alpha_invtlambda}.
%
%We assume that
%the system was initially equilibrated
%at a constant $\alpha_0$,
%such that the values of
%$\left\langle y_k^2(0) \right\rangle$
%can be computed from Eq. \eqref{eq:y2_eql},
%as well as Eq. \eqref{eq:t0_sinc}.
%
The total error is
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \frac
  {
    \Gamma_k \, \nu_k \,
    \left[
      \nu_k
      +
      (\nu_k - 1)
      \left(
        \frac{ t_0 } { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
    \right]
  }
  {
    (T + t_0) \, (2 \, \nu_k - 1)
  }
  .
\label{eq:error_invt}
\end{align}
%
%where
%$$
%\Err^f_k
%\equiv
%\left\langle
%  y_k^2
%\right\rangle_{ \alpha(t) }
%=
%\frac{  \Gamma_k \, \nu_k   }
%     {    2 \, (T + t_0)    }
%,
%$$
%is the equilibrium, or saturated, error of mode $k$
%at the terminal updating magnitude
%of $\alpha(T) = 1/[\lambda (T + t_0)]$.
